---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  message = FALSE,
  warning = FALSE
)
```

# OpenSDP Data

A project to generate realistic synthetic unit-level longitudinal education data. 

## Design Goals

1. Generate synthetic education data that is realistic for use by analysts across the education sector. 
 Realistic means messy, and reflective of the general pattern of relationships found in the U.S. education 
 sector. 
2. Synthetic data should be able to be generated on-demand and responsive to inputs from the user. These 
 inputs should allow the user to configure the process to produce data that resembles the patterns of data 
 in their agency. 
3. The package should be modular and extendable allowing new data topics to be generated as needed so synthetic 
 data coverage can grow. 

# Get Started

To use `OpenSDP.data`, follow the instructions below:

## Install Package

```{r eval=FALSE}
devtools::install_github("strategicdataproject/OpenSDP.data")
```

## Make some data

Using the `wakefield` package we can generate a simple set of demographic data. 

```{r, message=FALSE}
library(OpenSDP.data)
library(magrittr)
library(wakefield)
library(lubridate)
set.seed(612)

demog_master <- r_data_frame(n = 500, 
                             id(random = TRUE), 
                             sex, 
                             # dob, set range of years available for birth
                             dob(start = Sys.Date() - 365 * 25, 
                                 k = 365 * 8, by = "1 days"), 
                             race(x = c("White", "Hispanic or Latino Ethnicity", 
                                        "Black or African American", 
                                        "Asian", "American Indian or Alaska Native", 
                                        "Native Hawaiian or Other Pacific Islander", 
                                        "Demographic Race Two or More Races"), 
                                  prob = c(0.637, 0.163, 0.122, 0.047, .007, .0015, .021)))

head(demog_master)
```

Next, let's break the "Race" variable into a series of indicator variables.

```{r}
demog_master %<>% make_inds("Race")
demog_master %<>% mutate_at(5:11, 
                        funs(recode(., `0` = "No", `1` = "Yes")))
head(demog_master[, 4:9])
```

Now, let's generate some variables conditional on `race`. To do this we build a 
list that defines the distribution of this new variable for each category of 
the factor level. 

```{r}
# List of conditional probabilties
ses_list <- list("White" = list(f = rnorm, 
                                pars = list(mean = 0.3, sd = 1.1)), 
                 "Hispanic or Latino Ethnicity" = list(f = rnorm, 
                                pars = list(mean = -0.1, sd = 0.9)),
                 "Black or African American" = list(f = rnorm, 
                                pars = list(mean = -0.2, sd = 1.2)), 
                    "Asian" = list(f = rnorm, 
                                pars = list(mean = 0.23, sd = 1.2)), 
                 "Demographic Race Two or More Races" = list(f = rnorm, 
                                pars = list(mean = 0.0, sd = 1)), 
                 "American Indian or Alaska Native" = list(f = rnorm, 
                                pars = list(mean = -0.2, sd = 1)), 
                    "Other" = list(f = rnorm, 
                                pars = list(mean = 0, sd = 1)),
                 "Native Hawaiian or Other Pacific Islander" = list(f = rnorm, 
                                pars = list(mean = 0, sd = 1))
                    )

ses_list_b <- list("White" = list(f = rbinom, 
                                pars = list(size = 1, prob = 0.4)), 
                 "Hispanic or Latino Ethnicity" = list(f = rbinom, 
                              pars = list(size = 1, prob = 0.6)),
                 "Black or African American" = list(f = rbinom, 
                              pars = list(size = 1, prob = 0.65)), 
                 "Asian" = list(f = rbinom, 
                                pars = list(size = 1, prob = 0.375)), 
                 "Demographic Race Two or More Races" = list(f = rbinom, 
                                pars = list(size = 1, prob = 0.4)), 
                 "American Indian or Alaska Native" = list(f = rbinom, 
                              pars = list(size = 1, prob = 0.4)), 
                 "Other" = list(f = rbinom, 
                                pars = list(size = 1, prob = 0.4)),
                 "Native Hawaiian or Other Pacific Islander" = list(f = rbinom, 
                                  pars = list(size = 1, prob = 0.4))
)

# Note that cond_prob returns the whole data object
demog_master <- as.data.frame(demog_master)
demog_master <- cond_prob(demog_master, factor = "Race", 
                 newvar = "ses", prob_list = ses_list_b)

head(demog_master)
```

Now we have basic individual demographics, let's add annual attributes. 

```{r}
## Generate student-year data
minyear <- 1997
maxyear <- 2016
stu_year <- vector(mode = "list", nrow(demog_master))

# Make a list of dataframes, one for each student, for each year
for(i in 1:nrow(demog_master)){
  tmp <- expand.grid.df(demog_master[i, c(1, 3)], 
                        data.frame(year = 1:12))
  
  tmp$year <- lubridate::year(tmp$DOB + (tmp$year + 4) * 365)
  tmp$year - lubridate::year(tmp$DOB)
  stu_year[[i]] <- tmp; rm(tmp)
}

stu_year <- bind_rows(stu_year) %>% as.data.frame()
stu_year$age <- age_calc(dob = stu_year$DOB, 
                         enddate = as.Date(paste0(stu_year$year, "-09-21")),
                         units = "years", precise = TRUE)

head(stu_year)
```

Create an ELL indicator:

```{r}
# Create ELL
### Initial
## Identify first enrollment period for a student
## Look up probability based on age/race of being ELL
## Assign student to ELL status or not in first year

### Longitudinal
## If a student is not ELL, give a very very low probability of being ELL in the 
## future (.0001 in t + 1, .000001 in t+n)
## If a student is ELL, define a function for probability of exiting ELL status

stu_first <- stu_year %>% group_by(ID) %>% 
  mutate(flag = if_else(age == min(age), 1, 0)) %>% 
  filter(flag == 1) %>% select(-flag) %>% as.data.frame()

stu_first <- inner_join(stu_first, demog_master[, c("ID", "Race")])
stu_first$age <- round(stu_first$age, 0)
stu_first$race <- map_CEDS(stu_first$Race)
stu_first$ell <- assign_baseline(baseline = "ell", data = stu_first)
head(stu_first)
```


## Package Dependencies

- `dplyr`
- `lubridate`
- [wakefield](https://www.github.com/trinker/wakefield)


## OpenSDP

`OpenSDP.data` is part of the OpenSDP project. 
