---
title: "SDP College Going Analyze Example"
author: "Jared E. Knowles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CG Analyze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "tools/figs/README-",
  message = FALSE,
  warning = FALSE,
  fig.width =8, fig.height = 6.25
)
```

## Open SDP Data

First, we need to generate data suitable to conducting a college-going analysis. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(OpenSDP.data)
library(ggplot2)
library(tidyr)
library(scales)
library(magrittr)

simouts <- simpop(nstu = 35000, seed = 488234, 
                  control = sim_control(nschls = 12L, minyear=1996, 
                                        n_cohorts = 3,
                                        maxyear=2017)) 
cgdata <- sdp_cleaner(simouts)

dupes <- cg_data %>% group_by(sid) %>% mutate(dupes = n()) %>% filter(dupes > 1)
# 
# // Ninth grade cohorts you can observe persisting to the second year of college
# global chrt_ninth_begin_persist_yr2 = 2008
# global chrt_ninth_end_persist_yr2 = 2010
#  
# // Ninth grade cohorts you can observe graduating high school on time
# global chrt_ninth_begin_grad = 2013
# global chrt_ninth_end_grad = 2015
#  
# // Ninth grade cohorts you can observe graduating high school one year late
# global chrt_ninth_begin_grad_late = 2012
# global chrt_ninth_end_grad_late = 2014
#  
# // High school graduation cohorts you can observe enrolling in college the fall after graduation
# global chrt_grad_begin = 2009
# global chrt_grad_end = 2011
#  
# // High school graduation cohorts you can observe enrolling in college two years after hs graduation
# global chrt_grad_begin_delayed = 2008
# global chrt_grad_end_delayed = 2010

```


## Attainment

### Overall Progression

```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2004 &
                   chrt_ninth <= 2008) %>% 
  filter(!is.na(ontime))
plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)

agencyData <- plotdf %>%
  summarize(grad = mean(grad),
            seamless_transitioners_any = mean(seamless_transitioners_any, na.rm=TRUE),
            second_year_persisters = mean(second_year_persisters, na.rm=TRUE),
            N = n())
agencyData$school_name <- "AGENCY AVERAGE"
# // 2. Calculate the mean of each outcome variable by first high school attended
schoolData <- plotdf %>% group_by(first_hs_code) %>%
  summarize(grad = mean(grad),
            seamless_transitioners_any = mean(seamless_transitioners_any,
                                              na.rm=TRUE),
            second_year_persisters = mean(second_year_persisters, na.rm=TRUE),
            N = n())
## high school attended
names(schoolData)[1] <- "school_name"
# // 3. Identify the agency maximum values for each of the three outcome variables
maxSchool <- schoolData %>% summarize_all(.funs = funs("max"))
maxSchool$school_name <- "AGENCY MAX HS"
# // 4. Identify the agency minimum values for each of the three outcome variables
minSchool <- schoolData %>% summarize_all(.funs = funs("min"))
minSchool$school_name <- "AGENCY MIN HS"
# // 5. Append the three tempfiles to the school-level file loaded into R
schoolData <- bind_rows(schoolData, agencyData,
minSchool, maxSchool)
rm(agencyData, minSchool, maxSchool)
# // Step 6: Prepare to graph the results
library(tidyr)
schoolData$cohort <- 1
schoolData <- schoolData %>% gather(key = outcome,
                                    value = measure, -N, -school_name)
schoolData$subset <- grepl("AGENCY", schoolData$school_name)
library(ggplot2)
library(scales)
schoolData$outcome[schoolData$outcome == "cohort"] <- "Ninth Graders"
schoolData$outcome[schoolData$outcome == "grad"] <- "On-time Graduates"
schoolData$outcome[schoolData$outcome == "seamless_transitioners_any"] <-
  "Seamless College Transitioner"
schoolData$outcome[schoolData$outcome == "second_year_persisters"] <-
  "Second Year Persisters"

```

```{r}
ggplot(schoolData[schoolData$subset,],
aes(x = outcome, y = measure, group = school_name,
color = school_name, linetype = school_name)) +
geom_line(size = 1.1) + geom_point(aes(group = 1), color = I("black")) +
geom_text(aes(label = round(measure * 100, 1)), vjust = -0.8, hjust = -0.25,
color = I("black")) +
scale_y_continuous(limits = c(0, 1), label = percent) +
theme_bw() + theme(legend.position = c(0.825, 0.825)) +
guides(color = guide_legend("", keywidth = 6,
label.theme = element_text(face = "bold",
size = 8,
angle = 0)),
linetype = "none") +
labs(y = "Percent of Ninth Graders",
title = "Student Progression from 9th Grade Through College",
subtitle = "Agency Average", x = "",
caption = paste0("Sample: 2004-2005 Agency first-time ninth graders. \n",
"Postsecondary enrollment outcomes from NSC matched records. \n",
"All other data from Agency administrative records."))
```

### Progression by Race and Ethnicity

```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2004 &
                   chrt_ninth <= 2008)%>% 
  filter(!is.na(ontime))
plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)

# // Step 3: Create agency-level average outcomes
progressRace <- plotdf %>% group_by(race_ethnicity) %>%
  summarize(grad = mean(grad),
            seamless_transitioners_any = mean(seamless_transitioners_any, na.rm=TRUE),
            second_year_persisters = mean(second_year_persisters, na.rm=TRUE), N = n())
# // Step 4: Reformat the data for plotting
progressRace$cohort <- 1
progressRace <- progressRace %>% gather(key = outcome,
                                        value = measure, -N, -race_ethnicity)
# // Step 5: Recode variables for plot-friendly labels
progressRace$outcome[progressRace$outcome == "cohort"] <- "Ninth Graders"
progressRace$outcome[progressRace$outcome == "grad"] <- "On-time Graduates"
progressRace$outcome[progressRace$outcome == "seameless_transitioners_any"] <-
  "Seamless College Transitioner"
progressRace$outcome[progressRace$outcome == "second_year_persisters"] <-
  "Second Year Persisters"
progressRace$subset <- ifelse(progressRace$race_ethnicity %in% 
              c("Black or African American", "White", "Asian", "Hispanic or Latino Ethnicity"),
                              TRUE, FALSE)

```

```{r}
ggplot(progressRace[progressRace$subset,],
       aes(x = outcome, y = measure, group = race_ethnicity,
           color = race_ethnicity, linetype = race_ethnicity)) +
  geom_line(size = 1.1) + geom_point(aes(group = 1), color = I("black")) +
  geom_text(aes(label = round(measure * 100, 1)), vjust = -0.8, 
            hjust = -0.25, color = I("black")) +
scale_y_continuous(limits = c(0, 1), label = percent) +
theme_bw() + theme(legend.position = c(0.825, 0.825)) +
guides(color = guide_legend("", keywidth = 6,
                            label.theme =element_text(face = "bold", size = 8,
                                                      angle = 0)), linetype = "none") +
labs(y = "Percent of Ninth Graders",
title = "Student Progression from 9th Grade Through College",
subtitle = "By Student Race/Ethnicity", x = "",
caption = paste0("Sample: 2004-2005 Agency first-time ninth graders. \n",
"Postsecondary enrollment outcomes from NSC matched records. \n",
"All other data from Agency administrative records."))
```

### Progression by Race/Ethnicity by FRL


```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2004 &
                   chrt_ninth <= 2008)%>% 
  filter(!is.na(ontime))
plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)
plotdf <- plotdf %>% filter(frpl_ever_hs == 0)

progressRaceFRL <- plotdf %>% group_by(race_ethnicity) %>%
  summarize(grad = mean(grad),
            seameless_transitioners_any = mean(seamless_transitioners_any, na.rm=TRUE),
            second_year_persisters = mean(second_year_persisters, na.rm=TRUE),
N = n())
# // Step 5: Reformat the data file so that one variable contains all the
# outcomes of interest
progressRaceFRL %<>% filter(N >= 20)
# // Step 6: Prepare to graph the results
## Reshape the data
progressRaceFRL$cohort <- 1
progressRaceFRL <- progressRaceFRL %>% 
  gather(key = outcome,value = measure, -N, -race_ethnicity)

## Recode the variables for plot friendly labels
# // Step 5: Recode variables for plot-friendly labels
progressRaceFRL$outcome[progressRaceFRL$outcome == "cohort"] <- "Ninth Graders"
progressRaceFRL$outcome[progressRaceFRL$outcome == "grad"] <- "On-time Graduates"
progressRaceFRL$outcome[progressRaceFRL$outcome == "seameless_transitioners_any"] <-
  "Seamless College Transitioner"
progressRaceFRL$outcome[progressRaceFRL$outcome == "second_year_persisters"] <-
  "Second Year Persisters"
progressRaceFRL$subset <- ifelse(progressRaceFRL$race_ethnicity %in% 
              c("Black or African American", "White", "Asian", "Hispanic or Latino Ethnicity"),
                              TRUE, FALSE)
```



```{r}
ggplot(
  progressRaceFRL[progressRaceFRL$subset, ],
  aes(
  x = outcome,
  y = measure,
  group = race_ethnicity,
  color = race_ethnicity,
  linetype = race_ethnicity
  )
  ) +
  geom_line(size = 1.1) + geom_point(aes(group = 1), color = I("black")) +
  geom_text(
  aes(label = round(measure * 100, 1)),
  vjust = -0.8,
  hjust = -0.25,
  color = I("black")
  ) +
  scale_y_continuous(limits = c(0, 1), label = percent) +
  theme_bw() + theme(legend.position = c(0.825, 0.825)) +
  guides(color = guide_legend(
  "",
  keywidth = 6,
  label.theme = element_text(face = "bold", size = 8, angle = 0)
  ),
  linetype = "none") +
  labs(
  y = "Percent of Ninth Graders",
  title = "Student Progression from 9th Grade Through College",
  subtitle = paste0(
  c(
  "Among Students Qualifying for Free or Reduced Price Lunch \n",
  "By Student Race/Ethnicity"
  )
  ),
  x = "",
  caption = paste0(
  "Sample: 2004-2005 Agency first-time ninth graders. \n",
  "Postsecondary enrollment outcomes from NSC matched records.\n",
  "All other data from Agency administrative records."
  )
  )
```

### Progression by On-Track Status

```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2004 &
                   chrt_ninth <= 2008)

plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)

# // Step 3: Generate on track indicators that take into account students’ GPAs
# upon completion of their first year in high school
plotdf$ot <- NA
plotdf$ot[plotdf$ontrack_yr1 == 0] <- "Off-Track to Graduate"
# Check for correctness
plotdf$ot[plotdf$ontrack_yr1 == 1 & plotdf$cum_gpa_yr1 < 3 &
            !is.na(plotdf$cum_gpa_yr1)] <- "On-Track to Graduate, GPA < 3.0"
plotdf$ot[plotdf$ontrack_yr1 == 1 & plotdf$cum_gpa_yr1 >= 3 &
            !is.na(plotdf$cum_gpa_yr1)] <- "On-Track to Graduate, GPA >= 3.0"
# // Step 4: Calculate aggregates for the Agency by on track status
progressTrack <- plotdf %>% group_by(ot) %>%
  summarize(grad = mean(grad),
    seamless_transitioners_any = mean(seamless_transitioners_any, na.rm=TRUE),
    second_year_persisters = mean(second_year_persisters, na.rm=TRUE),
    N = n())

# of interest
progressTrack$cohort <- 1
progressTrack <- progressTrack %>% gather(key = outcome,
value = measure, -N, -ot)
progressTrack$outcome[progressTrack$outcome == "cohort"] <- "Ninth Graders"
progressTrack$outcome[progressTrack$outcome == "grad"] <- "On-time Graduates"
progressTrack$outcome[progressTrack$outcome == "seamless_transitioners_any"] <-
"Seamless College Transitioner"
progressTrack$outcome[progressTrack$outcome == "second_year_persisters"] <-
"Second Year Persisters"

```

```{r}
ann_txt <- data.frame(outcome = rep("Second Year Persisters", 3),
measure = c(0.22, 0.55, 0.85),
textlabel = c("Off-Track \nto Graduate",
"On-Track to Graduate,\n GPA < 3.0",
"On-Track to Graduate,\n GPA >= 3.0"))
ann_txt$ot <- ann_txt$textlabel
ggplot(progressTrack,
aes(x = outcome, y = measure, group = ot,
color = ot, linetype = ot)) +
geom_line(size = 1.1) + geom_point(aes(group = 1), color = I("black")) +
geom_text(aes(label = round(measure * 100, 1)), vjust = -0.8, hjust = -0.25,
color = I("black")) +
geom_text(data = ann_txt, aes(label = textlabel)) +
scale_y_continuous(limits = c(0, 1), label = percent) +
theme_bw() + theme(legend.position = c(0.825, 0.825)) +
scale_color_brewer(type = "qual", palette = 2) +
guides(color = "none",
linetype = "none") +
labs(y = "Percent of Ninth Graders",
title = "Student Progression from 9th Grade Through College",
subtitle = "By Course Credits and GPA after First High School Year", x = "",
caption = paste0("Sample: 2004-2005 Agency first-time ninth graders. \n",
"Postsecondary enrollment outcomes from NSC matched records. \n",
"All other data from Agency administrative records."))
```

## Ninth to Tenth Grade Transition by On-Track Status


### Proportion of Students On-Track by High School

```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2005 &
                   chrt_ninth <= 2008) %>% 
  filter(!is.na(ontime))

plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)

plotdf$ot <- NA
plotdf$ot[plotdf$ontrack_yr1 == 0] <- "Off-Track to Graduate"
plotdf$ot[plotdf$ontrack_yr1 == 1 & plotdf$cum_gpa_yr1 < 3 &
            !is.na(plotdf$cum_gpa_yr1)] <- "On-Track to Graduate, GPA < 3.0"
plotdf$ot[plotdf$ontrack_yr1 == 1 & plotdf$cum_gpa_yr1 >= 3 &
            !is.na(plotdf$cum_gpa_yr1)] <- "On-Track to Graduate, GPA >= 3.0"

progressBars <- bind_rows(
  plotdf %>% group_by(ot) %>% tally() %>% ungroup %>%
    mutate(count = sum(n), first_hs_code = "Agency Average"),
  plotdf %>% group_by(first_hs_code, ot) %>% tally() %>% ungroup %>%
    group_by(first_hs_code) %>%
    mutate(count = sum(n))
)

# replace first_hs_name = subinstr(first_hs_name, " High School", "", .)
# progressBars$first_hs_name <- gsub(" High School", "", progressBars$first_hs_name)
# // Step 5: For students who are off-track upon completion of their first year
# of high school, convert the values to be negative for ease of
# visualization in the graph
progressBars$n[progressBars$ot == "Off-Track to Graduate"] <- 
  -progressBars$n[progressBars$ot == "Off-Track to Graduate"]

```

```{r}
# // Step 6: Plot
ggplot(progressBars, aes(x = reorder(first_hs_code, n/count),
                         y = n/count, group = ot)) +
  geom_bar(aes(fill = ot), stat = 'identity', color = I("black")) +
  geom_text(aes(label = round(100* n/count, 0)),
            position = position_stack(vjust=0.3)) +
  theme_bw() +
  scale_y_continuous(limits = c(-0.8,1), label = percent,
                     name = "Percent of Ninth Graders",
                     breaks = seq(-0.8, 1, 0.2)) +
  scale_fill_brewer(name = "", type = "qual", palette = 6) +
  theme(axis.text.x = element_text(angle = 30, color = "black", vjust = 0.5),
        legend.position = c(0.15, 0.875)) +
  labs(title = "Proportion of Students On-Track to Graduate by School",
       subtitle = "End of Ninth Grade On-Track Status \n By High School", x = "",
       caption = paste0("Sample: 2004-2005 and 2005-20065 Agency first-time ninth
graders. \n",
"Postsecondary enrollment outcomes from NSC matched records. \n",
"All other data from Agency administrative records."))
```


### Ninth to Tenth Grade Transition by On-Track Status


## High School Graduation

### High School Completion Rates by School

```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2005 &
                   chrt_ninth <= 2008) %>% 
  filter(!is.na(ontime))

plotdf$grad <- ifelse(!is.na(plotdf$chrt_grad) & plotdf$ontime ==1, 1, 0)
plotdf$seamless_transitioners_any <- ifelse(plotdf$ps == 1, 1, 0)
plotdf$second_year_persisters <- as.numeric(plotdf$ps == 1 &
                                             plotdf$enroll_any_yr2 == 1)

schoolLevel <- bind_rows(
  plotdf %>% group_by(first_hs_code) %>%
    summarize(ontime_grad = mean(ontime, na.rm=TRUE),
              late_grad = mean(late, na.rm=TRUE),
              count = n()),
  plotdf %>% ungroup %>%
    summarize(first_hs_code = "Agency AVERAGE",
              ontime_grad = mean(ontime, na.rm=TRUE),
              late_grad = mean(late, na.rm=TRUE),
              count = n())
)
# // Step 3: Reshape the data wide
schoolLevel <- schoolLevel %>% gather(key = outcome,
                                      value = measure, -count, -first_hs_code)
schoolLevel$outcome[schoolLevel$outcome == "ontime_grad"] <- "On-Time HS Graduate"
schoolLevel$outcome[schoolLevel$outcome == "late_grad"] <- "Graduate in 4+ Years"
```

```{r}
ggplot(schoolLevel, aes(x = reorder(first_hs_code, measure), y = measure,
group = first_hs_code, fill = outcome)) +
  geom_bar(aes(fill = outcome), stat = 'identity', color = I("black")) +
  geom_text(aes(label = round(100 * measure, 0)),
            position = position_stack(vjust = 0.8)) +
  theme_bw() + theme(panel.grid = element_blank(), axis.ticks.x = element_blank()) +
  scale_y_continuous(limits = c(0, 1), label = percent, 
                     name = "Percent of Ninth Graders") + 
  scale_fill_brewer(name = "", type = "qual", palette = 7) +
  theme(axis.text.x = element_text(color = "black", angle = 30, vjust = 0.5),
        legend.position = c(0.15, 0.825)) +
  labs(title = "High School Graduation Rates by High School",
       x = "",
       caption = paste0("Sample: 2004-2005 Agency first-time ninth graders. \n",
                        "Data from Agency administrative records."))

```


### High School Completion Rates by Average 8th Grade Achievement


```{r}
plotdf <- filter(cg_data, chrt_ninth >= 2005 &
                   chrt_ninth <= 2008) %>% 
  filter(!is.na(ontime)) %>% filter(!is.na(test_math_8_std))


schoolLevel <- bind_rows(
plotdf %>% group_by(first_hs_code) %>%
summarize(ontime_grad = mean(ontime, na.rm=TRUE),
std_score = mean(test_math_8_std, na.rm=TRUE),
count = n()),
plotdf %>% ungroup %>%
summarize(first_hs_code = "Agency AVERAGE",
ontime_grad = mean(ontime, na.rm=TRUE),
std_score = mean(test_math_8_std, na.rm=TRUE),
count = n())
)



```

```{r}
ggplot(schoolLevel[schoolLevel$first_hs_code != "Agency AVERAGE", ],
       aes(x = std_score, y = ontime_grad)) +
  geom_vline(xintercept = as.numeric(schoolLevel[schoolLevel$first_hs_code ==
                                                   "Agency AVERAGE", "std_score"]),
             linetype = 4, color = I("goldenrod"), size = 1.1) +
  geom_hline(yintercept = as.numeric(schoolLevel[schoolLevel$first_hs_code ==
                                                   "Agency AVERAGE", "ontime_grad"]),
             linetype = 4, color = I("purple"), size = 1.1) +
  geom_point(size = I(2)) +
  theme_bw() + theme(panel.grid = element_blank()) +coord_cartesian() +
  annotate(geom = "text", x = -.85, y = 0.025,
           label = "Below average math scores & \n below average graduation rates",
           size = I(2.5)) +
  annotate(geom = "text", x = .85, y = 0.025,
           label = "Above average math scores & \n below average graduation rates",
           size = I(2.5)) +
  annotate(geom = "text", x = .85, y = 0.975,
           label = "Above average math scores & \n above average graduation rates",
           size = I(2.5)) +
  annotate(geom = "text", x = -.85, y = 0.975,
           label = "Below average math scores & \n above average graduation rates",
           size = I(2.5)) +
  annotate(geom = "text", x = .205, y = 0.025,
           label = "Agency Average \n Test Score",
           size = I(2.5), color = I("goldenrod")) +
  annotate(geom = "text", x = .85, y = 0.61,
           label = "Agency Average Graduation Rate",
           size = I(2.5)) +
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.2)) +
  scale_y_continuous(limits = c(0, 1), label = percent,
                     name = "Percent of Ninth Graders", breaks = seq(0, 1, 0.1)) +
  geom_text(aes(label = first_hs_code), nudge_y = 0.065, vjust = "top", size = I(4),
            nudge_x = 0.01) +
  labs(title = "High School Graduation Rates by High School",
       x = "Average 8th Grade Math Standardized Score",
       subtitle = "By Student Achievement Profile Upon High School Entry",
       caption = paste0("Sample: 2004-2005 through 2005-2006 Agency first-time ",
"ninth graders with eighth grade math test scores. \n",
"Data from Agency administrative records."))
```

