---
title: "Controlling the Simulation in OpenSDPsynthR"
author: "Jared E. Knowles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling the Data Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The `OpenSDPsynthR` package provides a set of functions to allow users to 
generate synthetic, but authentic, student-level data. Synthetic data is 
intended to make it easy to collaborate with analysts across the country 
tackling similar problems but using a shared vocabulary. The synthetic data 
will allow users to collaborate directly on code and analyses and verify that 
their analysis is working on synthetic data before translating it to live 
local data. 

This vignette explains how the package is structured so that it can be modified 
to meet the needs of users. 

```{r}
library(OpenSDPsynthR)
default_sim <- sim_control()
```

```{r}
names(default_sim)
```

There are over 25 user parameters that can be modified to control the simulation. 
The goal of these parameters is to allow the simulated student population to 
reflect a variety of possible educational environments ranging from small 
rural communities to large urban school districts. 

These parameters can have complex structures to allow for conditional and random 
generation of data. Parameters fall into four categories:

- **vectors:** a single list of parameters like school names, category names, or 
school IDs
- **conditional probability list:** an R list that contains a variable to group by, 
a function to generate data with, and a list of parameters for that function for 
each group in the grouping variable
- **outcome simulation parameters:** an R list of arguments to pass to the `simglm` 
function
- **outcome adjustments:** an R list of lists, with functions that modify a variable 
in an existing data set

### Vectors

The following vectors can be modified by the users:

- `nschls`: integer, number of high schools to assign students to
- `best_schl`: character, length 1, school ID for the highest performing school,
e.g. ("01")
- `race_groups`: character, length ?, names of racial subgroups to create in the 
simulation, defaults to US Census Groups
- `race_prob`: numeric, length = length(race_groups), proportion of population 
in each racial group
- `minyear`: integer, length 1, the first year of student data available
- `maxyear`: integer, length 1, the last year of student data available
- `n_cohorts`: integer, length 1, the number of graduation cohorts to create
- `school_names`: character, length = `nschls`, names of schools
- `assess_grades`: character, grade levels to simulate assessment scores for
- `postsec_names`: character, length = `n_postsec`, names of postsecondary schools
- `postsec_method`: character, length = 1, name of method to draw postsecondary 
schools from 

### Conditional Probability List

A conditional probability list is a list of lists in R. The `GROUPVARS` element 
specifies the grouping variable to conditionally assign probabilities. For 
example, if students are assigned gifted and talented status differently based 
on their sex, then this would specify `Sex`. The other elements of the list 
will be a separate list for each valid value of `Sex` -- in this case `Male` 
and `Female`. 

`Male` and `Female` are both lists that have two elements: `f` and `pars`. `f`
defines a function that is used to generate the variable, and `pars` contains 
all of the parameters for that function. 

```{r}
str(default_sim$gifted_list)
```

- `gifted_list`: a list defining how students are assigned to gifted and talented 
programs
- `iep_list`: a list defining how students are assigned to special education 
programs
- `ses_list`: a list deifning how students are assigned to free and reduced 
price lunch status
- `ell_list`: a list defining how students are assigned to English Language 
Learner status
- `ps_transfer_list`: a list defining the likelihood a student transfers 
postsecondary institutions


### Outcome Simulation Controls

Outcome simulation controls are lists with parameters to pass to the `simreg` 
function in the `simglm` package, which simulates hierarchical data and 
outcomes. 

Each of these simulations requires the user to specify:

- `fixed`: a RHS formula of the format `~ 1 + var1 + var2` defining the level 
1 variables for the simulation
- `random_var`: a numeric, length 1, specifying the variance in the second level 
- `cov_param`: a list, length of variables in `fixed` + 1 for the intercept, 
defines the function and parameters to generate the X values 
- `cor_vars`: a matrix of the variance between the X variables in `fixed`
- `fixed_param`: a vector of numerics, the lenth of `fixed` + 1, represent the 
beta coefficients
- `ngrps`: numeric, length of 1, number of second-level grouping terms
- `unbalanceRange`: numeric, length of 2, representing the minimum and maximum 
number of observations in each second-level cluster
- `type`: character, either "linear" or NULL

There are several of these parameters:

- `gpa_sim_parameters`: simulation parameters for the GPA simulation
- `grad_sim_parameters`: simulation parameters for high school graduation
- `ps_sim_parameters`: simulation parameters for postsecondary enrollment
- `assess_sim_par`: simulation parameters for student assessment data

### Outcome Simulation Adjustments

If we only rely on the simulation controls above, the data will be too predictable 
to be realistic, and structural inequalities along economic, racial, and gender 
lines will be underrepresented. To address this, it is possible to do post-simulation 
adjustments to introduce more variance to the outcomes. 

- `race_list`: 
- `perturb_race`: function, 
- `frl_list`: 
- `perturb_frl`: function, 

- `assessment_adjustment`: adjustments to the assessment score
- `grad_adjustment`: adjustments to the graduation probability
- `ps_adjustment`: adjustments to the postsecondary probability
- `gpa_adjustment`: adjustments to the grade point average


### Baselines

Currently there are two special parameters that are set based on baseline data 
built into the package. These are the initial grade distribution of students, 
and the initial program participation of students in `ell`, `iep`, and `frpl` 
programs. 

These set some of the simulation requirements, but others are set using the 
`baseline` function family.

```{r, eval=FALSE}
get_baseline("program")
get_baseline("grade")
```

Currently, baseline values cannot be modified by the user, but this will come in 
a future release. 

