---
title: "Controlling the Simulation in OpenSDP.data"
author: "Jared E. Knowles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling the Data Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

The `OpenSDP.data` package provides a set of functions to allow users to 
generate synthetic, but authentic, student-level data. Synthetic data is 
intended to make it easy to collaborate with analysts across the country 
tackling similar problems but using a shared vocabulary. The synthetic data 
will allow users to collaborate directly on code and analyses and verify that 
their analysis is working on synthetic data before translating it to live 
local data. 

This vignette explains how the package is structured so that it can be modified 
to meet the needs of users. 

```{r}
library(OpenSDP.data)
default_sim <- sim_control()
```

```{r}
default_sim
```


### Baselines

Currently there are two special parameters that are set based on baseline data 
built into the package. These are the initial grade distribution of students, 
and the initial program participation of students in `ell`, `iep`, and `frpl` 
programs. 

These set some of the simulation requirements, but others are set using the 
`baseline` function family.

```{r, eval=FALSE}
get_baseline("program")
get_baseline("grade")
```

## Longitudinal Indicators



