---
title: "Structuring the OpenSDPsynthR Package"
author: "Jared E. Knowles"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction


### Diagnostics

How do we know it worked? We can look at the patterns of ELL enrollment that 
are observed and see what patterns are the most common. To do this, let's compute 
the frequency of transition states observed per student. 

```{r genData}
library(OpenSDPsynthR)
simouts <- simpop(nstu = 1000, control = sim_control(nschls = 8L))
stu_year <- simouts$stu_year

```



```{r ellDiagnostic}
library(ggplot2)
library(tidyr)
plotdf <- stu_year %>% arrange(sid, year) %>% group_by(sid) %>% 
  do(tidy_sequence(.$ell, states = c(1, 0)))

plotdf$total <- rowSums(plotdf[, -1])
plotdf <- plotdf %>% gather(-sid, key = "Transition", value = "Count")

# plotdf %>% group_by(Transition) %>% filter(Transition != "total") %>%
#   summarize(sum(Count))

plotdf <- plotdf %>% filter(Transition != "total")  %>% 
  group_by(sid) %>% 
  mutate(total = sum(Count)) %>% 
  mutate(per = Count / total) %>% filter(Transition != "total")  %>% 
  separate(Transition, into = c("From", "To"), sep = "-") 

ggplot(plotdf, aes(Count)) + geom_histogram() + 
  scale_x_continuous(breaks = c(0:25)) + 
  facet_grid(From~To, labeller = label_both, switch = "y") + 
  theme_bw() + 
  labs(title = "Frequency of Transition States by Student - ELL", 
       y = "Count", x = "Times per Student State Observed")

```

Looking at this chart we can see that most students went from the No state to 
a No state -- as would be expected when there are few ELLs. 

Through this process we've gained students in the ELL status who were not 
initially ELL. Depending on our application this may not be desirable and we 
may want to modify the transition matrix to avoid this. Otherwise, later, 
this becomes an exercise in data cleaning. 

Two other visual diagnostics are below. 

```{r visualdiagnostics}
# Other plots

# ggplot(plotdf, aes(per)) + geom_density() + 
#   facet_grid(From ~ To, labeller = label_both, switch = "y") + 
#   theme_bw() + labs(title = "By Student Densities of Transitions")

# Heatmap
plotdf %>% group_by(From, To) %>% 
  summarise(Count = sum(Count)) %>% 
  ungroup %>% 
  mutate(total = sum(Count)) %>%
  mutate(per = Count/total) %>%
ggplot(aes(x = From, y = To, fill = per)) + 
  geom_tile(color= I("black")) + 
  geom_text(aes(label = round(per, digits = 2))) + 
  theme_minimal() +
  coord_cartesian() + labs(title = "Heatmap of ELL Transition States")

```


We can also do a comparative diagnostic. Given the relatively short length of 
our sequence per student, it will be hard to estimate fit from a short sequence. 


```{r testbyStudent}
# series <- stu_year$ell[stu_year$ID == "1705"]
# series <- stu_year$ell[stu_year$ID == "0001"]

test_fit <- function(series, expected){
  if(dim(table(series)) == 1){
    return(TRUE)
  } else {
  out <- fit_series(series, return = "fit", confidencelevel = 0.99, 
                    possibleStates = rownames(expected))
  low <- out$lowerEndpointMatrix < expected
  hi <- out$upperEndpointMatrix > expected
  return(all(low, hi))
  }
}

defaultFit <- sim_control()$ell_list$ALL$pars$tm

test_res <- stu_year %>% group_by(sid) %>% 
  summarize(fit_ok = test_fit(ell, expected = defaultFit))

table(test_res$fit_ok)
```


Let's look at co-occurrence of status over time. 

```{r CrossTabs}
# Look at by year patterns of relationships by student year
table(FRL = stu_year$frpl, GIFTED = stu_year$gifted)
table(FRL = stu_year$frpl, IEP = stu_year$iep)
table(GIFTED = stu_year$gifted, IEP = stu_year$iep)

```

Let's check polychoric correlations:

```{r}
gamma_GK(stu_year$gifted, stu_year$iep)
gamma_GK(stu_year$frpl, stu_year$iep)
gamma_GK(stu_year$frpl, stu_year$ell)
```


Finally, let's see who winds up "ever" in each category

```{r collapseEver}

test_df <- stu_year %>% group_by(sid) %>% 
  summarize(iep_ever = if_else(any(iep == 1), "Yes", "No"), 
            ell_ever = if_else(any(ell == 1), "Yes", "No"), 
            frpl_ever = if_else(any(frpl == 1), "Yes", "No"), 
            gifted_ever = if_else(any(gifted == 1), "Yes", "No"))

table(IEP_EVER = test_df$iep_ever)
table(ELL_EVER = test_df$ell_ever)
table(FRPL_EVER = test_df$frpl_ever)
table(GIFTED_EVER = test_df$gifted_ever)

```


## Assigning Schools and Outcomes

Students move through grades, schools, and outcomes. 

```{r, include=FALSE, eval=FALSE}
# Outcome assignment, outcomes are assigned in order
## sat_act
## ps_enroll

# TODO: Consider including diploma attainment...

out <- simpop(nstu = 1250, seed = 1241, sim_control(nschls = 9))
final_data <- sdp_cleaner(out)

ggplot(out$assessment, aes(x = age, y = math_ss, group = sid)) + 
  geom_line(alpha = I(0.2)) + 
  # geom_smooth(method = 'lm', se=FALSE, color = I("black"), alpha = I(0.2)) +
  facet_wrap(~schid) 


score_table <- assess %>% group_by(year, age) %>% 
  summarize(read_mean = mean(rdg_ss), 
            read_sd = sd(rdg_ss), 
            math_mean = mean(math_ss), 
            math_sd = sd(math_ss))

assess <- left_join(assess, score_table)
assess$math_std <- (assess$math_ss - assess$math_mean) / assess$math_sd
assess$read_std <- (assess$rdg_ss - assess$read_mean) / assess$read_sd
cor(assess$math_std, assess$read_std, use = "pairwise")

ggplot(assess, aes(x = age, y = math_std, group = sid)) +
  facet_wrap(~schid) + geom_line(alpha = I(0.2)) +
  geom_smooth(aes(group=1), se = FALSE)



ggplot(assess, aes(x = math_ss, y = rdg_ss)) + geom_point(alpha = I(0.2))
ggplot(assess, aes(x = math_std, y = read_std)) + geom_point(alpha = I(0.2))

idx <- sample(unique(assess$sid), 12)
ggplot(assess[assess$sid %in% idx, ], aes(x = age, y = math_std)) + 
  facet_wrap(~sid) + geom_line() + geom_smooth(method = 'lm', se=FALSE) + 
  geom_hline(yintercept = 0, color = I("red")) + geom_point()

ggplot(assess[assess$sid %in% idx, ], aes(x = age, y = math_ss)) + 
  facet_wrap(~sid) + geom_line() + geom_smooth(se=FALSE) 




g12_cohort <- out$stu_year[out$stu_year$grade == "12", ]
g12_cohort <- na.omit(g12_cohort)
g12_cohort <- left_join(g12_cohort, out$demog_master[, 1:4], by = "sid")
g12_cohort$male <- ifelse(g12_cohort$Sex == "Male", 1, 0)
hs_outcomes <- OpenSDPsynthR:::assign_hs_outcomes(g12_cohort, 
                                                 control = sim_control())

zzz <- out$hs_outcomes

dddff <- do.call(gen_outcome_model, ps_sim_parameters)




df <- sim_glm(fixed = fixed, random = random,
                  fixed_param = fixed_param, random_param = random_param,
                  random3 = NULL,
                  random_param3 = NULL,
                  cov_param = cov_param,
                  fact_vars = fact_vars, k = NULL,
                  n = ngrps, p = NULL,
                  cor_vars = cor_vars, data_str = "cross", unbal = TRUE,
                  unbalCont = unbalanceRange)
    mod <- glmer(update(fixed, "sim_data ~ . - math_ss + (1|clustID)"),
                 data = df, family = "binomial")

# out <- simpop(nstu = 400, seed = 32231, 
#           control = sim_control(nschls = 3L))

# Student	Student Year	Outcome
# sid	assessment	hs_diploma
# race_ethnicity	school_id	cum_gpa_final
# sex	on_track	sat_act
# frpl_ever	frpl	ps_enroll
# ell_ever	ell	dropout
# gifted_ever	gifted	transfer
# iep_ever	iep	disappear
# grade_level	still_enroll
    random <- ~ 1
  random_param <- list(random_var = random_var, rand_gen = "rnorm")

 library(simglm)
  
assess_sim_par <- list(
  fixed = ~ 1 + time + gifted + iep + frpl + ell + male,
  random = ~ 1 + time,
  random3 = ~ 1 + time,
  cor_vars = c(-0.276, -0.309, -0.046, -0.033,
              -0.03, -0.029, -0.003, 0.06, 0.007, 0.001),
  fixed_param = c(0.0024, 0.75, 0.10, -0.161388, -0.075, -0.056, 0.007),
  fact_vars = NULL,
  # intercept + any slopes in length
  random_param = list(random_var = c(0.2, 0.1), cor_vars = c(0.4), rand_gen = 'rnorm'),
  random_param3 = list(random_var = c(0.3, 0.025), rand_gen = 'rnorm'), # intercept + any slopes in length
  cov_param = list(
                 dist_fun = c("rbinom", "rbinom", "rbinom", "rbinom", "rbinom"),
                 var_type = rep("lvl1", 5), 
                 opts = list(
                   list(size = 1, prob = 0.1), 
                   list(size = 1, prob = 0.2), 
                   list(size = 1, prob = 0.45), 
                   list(size = 1, prob = 0.1), 
                   list(size = 1, prob = 0.52)
                  )
               ),
  unbalCont = c(2, 16),
  unbalCont3 = c(100, 800),
  unbal = TRUE, 
  # Total number of level 2 groups = k * n
  k = 15, # level 3 groups
  n = 200, # obs per group level 2 group
  p = 400, # obs per group?
  error_var = 1,
  with_err_gen = 'rnorm',
  lvl1_err_params = list(mean = 0, sd = 1),
  data_str = "long"
)
  
 assess_table <- do.call(sim_reg, assess_sim_par, quote = TRUE)
 # needs to be the length of all correlations between predictors
temp_three <- sim_reg(fixed = fixed, random = random, random3 = random3,
                      fixed_param = fixed_param, random_param = random_param,
                      random_param3 = random_param3, cov_param = cov_param,
                      fact_vars = fact_vars, k = k,n = n, p = p,
                      lvl1_err_params = lvl1_err_params,
                      error_var= error_var, with_err_gen = with_err_gen,
                      cor_vars = cor_vars, data_str = "long", unbal = TRUE,
                      unbalCont = unbalCont, unbalCont3 = unbalCont3)
 

library(ggplot2)
ggplot(temp_three, aes(x = time, y = sim_data, group = clustID)) +
  geom_line(alpha = I(0.2)) + facet_wrap(~clust3ID)

names(temp_three)[1:6] <- c("intercept", "age", "gifted", "iep", "frpl",
                       "ell")
names(temp_three)[14] <- "math_ss"
names(temp_three)[15:17] <- c("time", "sid", "schid")

ggplot(temp_three, aes(x = age, y = math_ss, group = sid)) +
  geom_line(alpha = I(0.2)) + facet_wrap(~schid)

#witihnID = time, nested w/in level 2

library(lme4)
proof <- lmer(math_ss ~ 1 + age + gifted +
                iep + frpl + ell +
                (1 + age | sid) +
                (1 | schid), data = temp_three)   

```


